Simple repression draft

### 2020-05-19 doc/section_02_means.tex
initial splitting of document into several files

\documentclass[12pt]{article}%{amsart}
\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
%\geometry{landscape}             % Activate for for rotated page geometry
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{epstopdf}
% \usepackage{caption}
\usepackage{amsmath}
% \usepackage{longtable}
% \usepackage{tabu}
% \usepackage{accents} %to get undertildes for vec & mat
\usepackage{color} % for colored text
% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command

% \newcommand\mytablefigwidth{0.35\textwidth}

% allows to use .ai files directly w/o resaving as pdf
\DeclareGraphicsRule{.ai}{pdf}{.ai}{}

% Handy math macros!
\newcommand{\vect}[1]{\vec{#1}}
\newcommand{\matr}[1]{\mathbf{#1}}
\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}

% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
% usage: provide 1 arg: \deriv{x} gives d/dx
% provide 2 args like so: \deriv[f]{x} gives df/dx
\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}

%%%%% Referencing macros %%%%%
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\tref}[1]{Table~\ref{#1}}
\newcommand{\eq}[1]{Eq.~(\ref{#1})}
\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}

%%%%%
\begin{document}

\title{Simple repression draft}

\maketitle

\section{Introduction}
\mmnote{Key ideas, no particular order, that I haven't written down before:
\begin{itemize}
\item Our goal is to build phenomenological models of input-output
functions of genetic circuits. More precisely, we want phenomenology
in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning, e.g., we don't want to coarse-grain
as far as Hill functions. Why not? We are motivated by studies
like~\cite{Razo-Mejia2020}, for which Hill functions are
insufficient. Such work requires a quantitative theory of how
biophysical changes at the molecular level propagate to
input-output functions at the genetic circuit level.
We want concepts, not mere facts.
In particular a key question is: at this level of
coarse graining, what microscopic details do I need to explicitly
model, and how do we figure that out? For example, do I need to
worry about all or even any of the steps that individual RNAPs go
through each time they make a transcript? Turning the question
around, can we see any imprint of those processes in the
available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling can complement each other
beautifully here.
\end{itemize}}

\fig{fig1:means_cartoons}A shows the qualitative picture of
simple repression. An operator, the binding site on the DNA for a
repressor protein, may be found occupied by a repressor, in which
case transcription is blocked from occurring. Or it may be found
unoccupied, in which case RNA polymerase (RNAP) may bind and
transcription can proceed.

The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
It is often imagined that when the repressor is bound to its operator,
RNAP is sterically blocked from binding to its promoter sequence.
Current evidence suggests this is sometimes, but not
always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
Rob??, showing locations of repressor binding sites far upstream
of promoter}, and it remains an interesting open question precisely
how a repressor bound far upstream is able to repress transription.
Suggestions include ``action-at-a-distance'' mediated by kinks in
the DNA, formed when the repressor is bound, that prevent RNAP binding.
Nevertheless, our modeling in this work is sufficiently
coarse-grained that we simply assume exclusive binding and need
not worry about such precise details.

\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple
repression, let us consider what we can learn from theory and
measurements of fold-change in mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells
and mean mRNA $\langle m\rangle$ is viewed as a function of
repressor copy number $R$.
The second
equality in~\ref{eq:fc_def} follows from assuming that the
translation efficiency, i.e., the number of proteins translated
per mRNA, is the same in both conditions.
In other words, we assume that mean protein level is proportional
to mean mRNA level, and that the proportionality constant is the
same in both conditions and therefore cancels out in the ratio.
This is reasonable since the cells in the two conditions are
identical except for the presence of the transcription factor,
and it is difficult to imagine the transcription factor having
any direct effect on translation.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{
\textbf{
An overview of the simple repression motif at the level of means.}
(A) depicts the qualitative biological picture of simple repression.
(B) shows a variety of possible mathematized cartoons of simple
repression, along with the effective parameter $\rho$
which subsumes all regulatory details
of the architecture excluding the repressor.
(C) shows the ``master curve'' to which all cartoons in (B) collapse.
}
\label{fig1:means_cartoons}
\end{figure}

@@

In principle, if $\Delta F_R$ can be measured to sufficient
precision, then deviations from $\rho=1$ become a
testable matter of experiment.
In practice, it is probably unrealistic to measure repressor
rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
(and hence $\Delta\epsilon_R$) precisely enough to detect the
expected tiny deviations from $\rho=1$.
We can estimate the
requisite precision in $\Delta F_R$ to resolve a given
$\Delta\rho$ by noting, since $\rho\approx1$, that
$\log(1+\Delta\rho)\approx \Delta\rho$, so
$\Delta(\Delta F_R) \approx \Delta\rho$.
Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
the high end of our range estimated above.
A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
barely 0.1 was achieved in the excellent measurements
of~\cite{Razo-Mejia2018}, so this requires a very difficult
determination of $\Delta F_R$ for a very crude determination of $\rho$,
which suggests, to put it lightly, this is not a
promising path to pursue experimentally.
It is doubtful that inference of repressor kinetic rates
would be any easier.\footnote{
\mmnote{
In fact the uncertainties quoted in Hammar et.\
al.~\cite{Hammar2014} from Elf's group are small enough to do
this calculation, but HJ \& I think their error bars are insanely
optimistic. Not sure I want to blatantly call them out on it though??}
}

@@

\section{Beyond means}
\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
Before we can tackle simple repression, we need an adequate
phenomenological model of constitutive expression.
The literature abounds with options from which we can choose,
and we show several potential kinetic models for constitutive promoters
in~\fig{fig:constit_cartoons}.
Let us consider the suitability of each model for our purposes in turn.

\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
\caption{\textbf{Constitutive promoter cartoon comparison.}
The left column depicts various plausible cartoons for constitutive promoters.
We also list literature references where models have been used previously.
In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
Model (2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}.
Model (3) is a more detailed generalization of model (2),
treating transcription initiation as a multi-step process
proceeding through closed and open complexes~\cite{Mitarai2015}.
Model (4) is somewhat analogous to (2) except with the precise nature
of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
Finally, model (5) can be viewed as a certain limit of model (4)
in which transcripts are produced in bursts,
and initiation of bursts is a Poisson process.
\mmnote{Haven't found model 5 formulated quite like this in the literature,
usually burstiness is handled like model 4, but I should probably search a bit more.}
The right column shows the Fano factor for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$
strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data
observed in~\cite{Jones2014} without introducing additional
assumptions and model complexity.
}
\label{fig:constit_cartoons}
\end{figure}

\subsubsection{Model 1 - Poisson promoter.}
This is the picture from Jones et.\ al.~\cite{Jones2014}, which
assumes that transcripts are produced as a Poisson process from a
single promoter state. This model insists that the ``true'' mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute deviations
from Fano = 1 to intensity variability in fluorescent spots, gene
copy number variation, and copy number fluctuations of the
transcription machinery, e.g., RNAP itself.
In this picture, the master equation makes no appearance,
and all the corrections to Poisson behavior are derived
as additive corrections to the Fano factor.
For disproving the ``universal noise curve'' from So et
al~\cite{So2011}, this picture was excellent. It is appealing in
its simplicity, with only two parameters, the initiation rate $r$
and degradation rate $\gamma$. Since $\gamma$ is independently
known from other experiments, and the mean mRNA copy number is
$r/\gamma$, $r$ is easily inferred from data.
In other words, the model is not excessively complex for the data at hand.
But for many interesting questions, for instance in the recent
work~\mmnote{cite Manuel's preprint once posted},
knowledge of means and variances alone is insufficient, and a
description of the full distribution of molecular counts is necessary.
For this we need a( slightly) more complex model than model 1
that would allow us to incorporate the non-Poissonian features of
constitutive promoters directly into a master equation formulation.

\subsubsection{Model 2 - Two-state promoter, RNAP bound or unbound.}
This model was considered in, e.g.,~\cite{Phillips2015a} and ~\cite{Phillips2019}.
Here transcription initiation proceeds from the bound to the unbound state,
reflecting the microscopic reality that an RNAP that has begun to elongate
a transcript is no longer available at the start site to begin another.
The problem with this picture is that the Fano factor is
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)},
\end{align}
which is always $<1$.
To make contact with the experimental reality of $\nu>1$,
we will have to do, at a minimum,
the same corrections to Poisson behavior as in model 1 above.
So while this model adds an appealing element of microscopic
reality, we are forced to reject it as the additional complexity
is unable to capture the phenomenology of interest.
Obviously the promoter state does in fact proceed through cycles
of RNAP binding, initiating, and elongating, but it seems that
the super-Poissonian noise in mRNA copy number we want to
model must be governed by other features of the system.

\subsubsection{Model 3 - Three-state promoter, multistep
transcription initiation and escape.}
How might we remedy the deficits of model 2?
It is known~\cite{DeHaseth1998} that once RNAP initially binds
the promoter region, a multitude of distinct steps occur
sequentially before RNAP finally escapes into the elongation
phase. Perhaps adding some of this mechanistic detail might
rescue model 2? The next simplest refinement of model 2 could
consider open complex formation and promoter escape as separate
steps rather than as a single effective step. In other words, we
construct model 3 by adding a single extra state to model 2, and
we will label the two RNAP-bound states as the closed and open
complexes, despite the true biochemical details certainly being
more complex. The authors of~\cite{Mitarai2015}, for instance,
considered this model, although they added an additional
repressor bound state and did not explicitly consider the limit
with no repressor that we analyze here. Again, our goal here is
not a complete accounting of all the relevant biochemical detail;
this is an exploratory search for the important features our
model needs to include.

Unfortunately, as the authors of~\cite{Mitarai2015} hint, this
model too has Fano factor $\nu<1$. This can be seen by starting
with their Eq.~A1 for the Fano factor of the analogous model with
repressor and taking the limit as repressor concentration goes to
zero. After substantial algebra, the result is
\begin{align}
\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + \gamma}
        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$
for notational tidiness. This is necessarily less than 1 for
arbitrary rate constants.

In fact, we suspect \textit{any} model in which transcription
proceeds through a multistep cycle must necessarily have $\nu<1$.
The intuitive argument compares the waiting time distribution to
traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple
exponentials, and intuitively should be more peaked with a
smaller fractional width than an exponential with the same mean.\footnote{
This can be made more precise with a result
from~\cite{Stewart2007}, who showed that the convolution of
multiple gamma distributions (of which the exponential
distribution is a special case) is to a very good approximation
also gamma distributed. Using their Eq.~2 for the distribution of
the convolution, with shape parameters set to 1 to
give exponential distributions, the total waiting time
distribution has a ratio of variance to squared mean 
$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
where the $k_i$ are the rates of the individual steps. Clearly
this is less than 1 and therefore the total waiting time
distribution is narrower than the corresponding exponential.} A
less disperse waiting time distribution means transcription
initations are more uniformly distributed in time relative to a
Poisson process. Hence the distribution of mRNA over a population
of cells should be less variable compared to Poisson, giving
$\nu<1$.\footnote{
This last step, while intuitive, can be argued by analogy to
photon statistics where antibunching gives rise to sub-Poissonian
noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
not expect they apply for our problem. Nevertheless we refrain
from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
be totally out of left field for most readers, any suggestions on
a more comprehensible reference??}
}

Regardless of the merits of this model in describing the noise
properties of constitutive transcription initation, it does not
capture the dominant source of noise we want to include in our
phenomenological model, so we must discard these details and
search elsewhere.

\subsubsection{Model 4 - Two-state promoter,
``active'' and ``inactive'' states.}
Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model
2, but the interpretation of the two states is changed. Rather
than explicitly viewing them as RNAP bound and unbound, we view
them as ``active'' and ``inactive,'' which are able and unable to
initiate transcripts, respectively. We are noncommittal as to the
microscopic details of these states.

One interpretation~\cite{Chong2014, Sevier2016} for the active
and inactive states is that it represents the promoter's
supercoiling state: transitions to the inactive state are caused
by accumulation of positive supercoiling, which inhibits
transcription, and transitions back to ``active'' are caused by
gyrase or other topoisomerases relieving the supercoiling. This
is an interesting possibility because it would mean the timescale
for promoter state transitions is driven by topoisomerase
kinetics, not by RNAP kinetics. From in vitro measurments, the
former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are
assumed to be governed by RNAP, which, assuming a copy number per
cell of order $10^3$, has a diffusion-limited association rate
$k_{on} \sim 10^2~\text{s}^{-1}$ to a target promoter. Combined
with known $K_D$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$. As we will show
below, however, there are some lingering puzzles with
interpreting this supercoiling interpretation, so we leave it as
a speculation and refrain from assigning definite physical
meaning to the two states in this model.

Intuitively one might expect that, since transcripts are produced
as a Poisson process only when the promoter is in one of the two
states in this model, transcription initiations should now be
``bunched,'' in constrast to the ``anti-bunching'' of models 2
and 3 above. One might further guess that this bunching would
lead to super-Poissonian noise in the mRNA distribution over a
population of cells. An honest calculation of the Fano factor produces
\begin{align}
\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\end{align}
which is strictly greater than 1, verifying the above intuition.
Note we have dropped the $P$ label on the promoter switching rates to
emphasize that these very likely do not represent kinetics of RNAP itself.
This calculation can also be sidestepped by noting that the model
is mathematically equivalent to the simple repression repression model
from~\cite{Jones2014}, with states and rates relabeled and reinterpreted.

How does this model compare to model 1 above?
In model 1, all non-Poisson features of the mRNA
distribution were handled as extrinsic corrections.
By contrast, here the 3 parameter model is used to fit the full
mRNA distribution as measured in mRNA FISH experiments.
In essense, all variability in the mRNA distribution is regarded
as ``intrinsic,'' arising either from stochastic initiation or
from switching between the two coarse-grained promoter states.
The advantage of this approach is that it fits neatly into the
master equation picture, and the parameters thus inferred can be used as
input for more complicated models with regulation by transcription factors.

While this seems promising, there is a major drawback for our purposes
which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
the statistical inference problem is nonidentifiable,
in the sense described in Section 4.3 of~\cite{Gelman2013},
meaning it is impossible to infer the
parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
Rather, only the ratio $r/k^-$ could be inferred. In that work,
the problem was worked around with an informative prior on the
ratio $k^-/k^+$. That approach is unlikely to work here, as,
recall, our entire goal in modeling constitutive expression is to
use it as the basis for a yet more complicated model, when we add
on repression. But adding more complexity to a model that is
already poorly identified is a fool's errand,
so we will explore one more potential model.

\subsubsection{Model 5 - One-state promoter with explicit bursts.}
This model is inspired by the failure mode of model 4.
The key observation above was that, as found
in~\cite{Razo-Mejia2020}, only two parameters, $k^+$ and the
ratio $r/k^-$, could be directly inferred from the FISH data
from~\cite{Jones2014}.
So let us take this seriously and imagine a model where these are
the only two model parameters. What would this model look like?

To develop some intuition, consider model 4 in the limit
$k^+ \ll k^- \lesssim r$, which is roughly satisfied by the
parameters inferred in~\cite{Razo-Mejia2020}.
In this limit, the system spends the majority of its time in the inactive state, occasionally becoming active and making a burst of transcripts.
This should call to mind the well-known phenomenon of transcriptional bursting, as reported in, e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some more cites here}.
Let us make this correspondence more precise.
The mean dwell time in the active state is $1/k^-$.
While in this state, transcripts are produced at a rate $r$ per
unit time. So on average, $r/k^-$ transcripts are produced before
the system switches to the inactive state. Once in the inactive
state, the system dwells there for an average time $1/k^+$ before
returning to the active state and repeating the process. $r/k^-$
resembles an average burst size, and $1/k^+$ resembles the time
interval between burst events. More precisely, $1/k^+$ is the
mean time between the end of one burst and the start of the next,
whereas $1/k^+ + 1/k^-$ would be the mean interval between the
start of two successive burst events, but in the limit $k^+ \ll
k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
that the waiting time between bursts is approximately
exponentially distributed, with mean set by the only timescale
left in the problem,
$1/k^+$.\footnote{
If instead it were the case that $k^+ \sim k^-$, then the waiting time
$1/k^+ + 1/k^-$ would have a peaked distribution, but this does
not appear to be the case for any of datasets
from~\cite{Jones2014}.}

Let us now verify this intuition with a precise derivation
to check that $r/k^-$ is in fact the mean burst size
and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state.
Transcripts are produced at a rate $r$ per unit time, so the
number of transcripts $n$ produced during $T$ fits exactly
the ``story'' for the Poisson distribution, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable, we actually want $P(n)$,
the dwell time distribution with no conditioning on $T$.
Basic rules of probability theory tell us we can write $P(n)$
in terms of $P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$,
which is exponentially distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows
$\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain
$n \in \{0, 1, 2, \dots\}$
(this is one of two common conventions for the geometric distribution).
The mean of the geometric distribution, with this convention, is
\begin{align}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{align}
exactly as we guessed intuitively above.

So in taking the limit $r,k^-\rightarrow\infty$, $r/k^-\equiv b$,
we obtain a model which effectively has only a single promoter state,
which initiates bursts at rate $k^+$
(transitions to the active state, in the model 4 picture).
The master equation for mRNA copy number $m$ takes the form
\begin{align}
\begin{split}
\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
         - \sum_{j=m+1}^\infty k^+ p(m,t) Geom(j-m;b),
\end{split}
\end{align}
where $Geom(n;b)$ is the geometric distribution with mean~$b$, i.e.,
$Geom(n;b) = \frac{1}{1+b}\left(\frac{b}{1+b}\right)^n$
(with domain over nonnegative integers as above).
The first two terms are the usual mRNA degradation terms.
The third term enumerates all ways the system can produce
a burst of transcripts and arrive at copy number $m$,
given that it had copy number $j$ before the burst.
The fourth term allows the system to start with copy number $m$,
then produce a burst and end with copy number $j$.
In fact this last sum has trivial $j$ dependence and
simply enforces normalization of the geometric distribution.
Carrying it out we have
\begin{equation}
\begin{split}
\deriv{t}{p(m,t)} = & (m+1)\gamma p(m+1,t) - m\gamma p(m,t) \\
        &+ \sum_{j=0}^{m-1} k^+ p(j,t) Geom(m-j;b)
            - k^+ p(m,t),
\end{split}
\end{equation}
This improves on model 4 in that now the parameters are easily inferred,
as we will see later, and have clean interpretations.
The non-Poissonian features are attributed to the emprically
well-established phenomenological picture of bursty transcription.

The big approximation in going from model 4 to 5 is that a burst
is produced instantaneously rather than over a finite time.
If the true burst duration is not short compared to transcription factor
kinetic timescales, this could be a problem in that mean burst size
in the presense and absence of repressors could change,
rendering parameter inferences from the constitutive case inappropriate.
Let us make some simple estimates of this.

Consider the time delay between the first and final RNAPs
in a burst initiating transcription
(\textit{not} the time to complete transcripts,
which potentially could be much longer.)
If this timescale is short compared to the typical
search timescale of transcription factors, then all is well.
The estimates from deHaseth et.\ al.~\cite{DeHaseth1998}
put RNAP's diffusion-limited on rate around
$\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$
and polymerase loading as high as $1~\text{s}^{-1}$.
Then for reasonable burst sizes of $<10$, it is reasonable to guess that
bursts might finish initiating on a timescale of tens of seconds or less
(with another 30-60 sec to finish elongation, but that does not matter here).
A transcription factor with typical copy number of order 10 (or less)
would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$.
Higher copy number TFs tend to have many binding sites over the genome,
which should serve to pull them out of circulation and keep their
effective association rates from rising too large.
Therefore, there is \textit{perhaps} a timescale separation
possible between transcription factor association rates and burst
durations, but this assumption could very well break down, so we
will have to keep it in mind when we infer repressor rates from
the Jones et.\ al.\ FISH data later.

In reflecting on these 5 models, the reader may feel that
exploring a multitude of potential models just to return to a
very minimal phenomenological model of bursty transcription may
seem highly pedantic. But the purpose of the exercise was to
examine a host of models from the literature and understand why
they are insufficient, one way or another, for our purposes.
Along the way we have learned that the detailed kinetics of RNAP
binding and initiating transcription are probably irrelevant for
setting the population distribution of mRNA.
The timescales are simply too fast, and as we will see later
in~\fig{fig:constit_post}, the noise seems to be governed by
slower timescales. Perhaps in hindsight this is not surprising:
intuitively, the degradation rate $\gamma$ sets the fundamental
timescale for mRNA dynamics, and any other processes that
substantially modulate the mRNA distribution should not differ
from $\gamma$ by orders of magnitude.

\section{Finding the ``right'' model: Bayesian parameter inference and model selection}
\subsection{Parameter inference for constitutive promoters}

From consideration of Fano factors in the previous section, we
suspect that model 5 in~\fig{fig:constit_cartoons},
a one-state bursty model of constitutive promoters,
achieves a Goldilocks level of complexity.
Does this stand up to closer scrutiny, namely, comparison to
full mRNA distributions rather than simply their moments?
We will test this thoroughly on all the data
from constitutive promoters from Jones et.\ al.~\cite{Jones2014}.

It will be instructive, however, to first consider the
Poisson promoter, model 1 in~\fig{fig:constit_cartoons}.
This is a convenient proxy for models 2 and 3, whose exact
steady-state mRNA distributions would be tedious to calculate
and we will not explicitly consider.
But since they have Fano factors $\nu\le 1$, they can
fare no better than the simple Poisson model.
We will also not explicitly consider model 4 from~\fig{fig:constit_cartoons}
since it was already thoroughly analyzed in~\cite{Razo-Mejia2020},
and since model 5 can be viewed as a special case of it.

In performing parameter inference on the FISH data
of~\cite{Jones2014}, building a generative model is
straightforward. The data are single-cell mRNA counts, and we
assume different cells to be independent and identically
distributed from the underlying generative model, which is just
the steady-state mRNA distribution computed from the master
equation of the underlying model.

\subsubsection{Model 1: Poisson promoter}
\mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
For this model the master equation of interest
is~\eq{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
\end{equation}
whose steady-state distribution is Poisson, given by
\begin{equation}
p_{m,U} = \frac{\lambda^m e^{-\lambda}}{m!}
\end{equation}
with $\lambda=r/\gamma$. This steady state distribution defines our
likelihood $p(m\mid\lambda)$, i.e.,
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
\label{eq:poisson_inference010}
\end{equation}
is the probability of observing
a single cell with $m$ mRNAs given a known value of $\lambda$.

With our likelihood in hand, we can invoke Bayes theorem to write down
the posterior distribution
on our model parameter $\lambda$, given by
\begin{equation}
p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
\end{equation}
where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts
over a population of $N$ identical cells.
As is standard we have neglected the factor $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as
a normalization factor.
Since we assume each cell's mRNA count is independent of others,
the likelihood is simply a product of the
single cell likelihoods given by \eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}
This is often notated simply as
\begin{equation}
D\mid\lambda \sim \text{Poisson}(\lambda),
\end{equation}
which is read as the data $D$, conditioned on a value of
$\lambda$, is Poisson distributed with mean $\lambda$.
The $\sim$ operator literally denotes ``is distributed according to.''

To proceed we need to specify a prior. In this case we are extremely data-rich,
as the dataset from Jones et.\ al~\cite{Jones2014} has of order
1000-3000 single-cell measurements for each promoter,
so our choice of prior matters little here,
as long as it does not exclude the true value of $\lambda$.
A convenient choice for our Poisson likelihood is its conjugate prior,
the gamma distribution. Although they only exist for
certain special likelihoods, conjugate priors have the nice
feature that the posterior takes the same functional form as the
prior, with updated parameters. This makes calculations
analytically tractable and also offers a nice interpretation of
the inference procedure as updating our knowledge about the model
parameters.

Putting a gamma prior on $\lambda$ introduces two new parameters
$\alpha$ and $\beta$ which parametrize the gamma distribution
itself, which we use to encode the range of $\lambda$ values we
view as reasonable. Recall $\lambda$ is the mean steady-state
mRNA count per cell, which \textit{a priori} could plausibly be
anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$
achieve this, since the gamma distribution is strictly positive
with mean $\alpha/\beta$ and standard deviation
$\sqrt{\alpha}/\beta$. To be explicit, then, our prior is
\begin{equation}
\lambda\mid\alpha,\beta \sim \text{Gamma}(\alpha, \beta)
\end{equation}

As an aside, note that if we did not know that our prior was a
conjugate prior, we could still write down our posterior
distribution from its definition as
\begin{equation}
p(\lambda\mid D,\alpha,\beta)
\propto p(D\mid\lambda) p(\lambda \mid\alpha,\beta)
\propto \left(\prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}\right)
        \frac{\beta}{\Gamma(\alpha)}(\beta\lambda)^{\alpha-1} e^{-\beta\lambda}
.
\end{equation}
Without foreknowledge that this in fact reduces to a gamma distribution,
this expression might appear rather inscrutable. When conjugate
priors are unavailable for the likelihood of interest - which is
almost always the case for models with $>1$ model parameter -
this inscrutability is the norm, and making sense of posteriors
analytically is almost always impossible. Fortunately, MCMC
sampling provides us a powerful method of constructing posteriors
numerically which we will make use of extensively.

>>>>

% !TEX root = ./busty_transcription.tex
\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple repression, let
us consider what we can learn from theory and measurements of fold-change in
mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells and mean mRNA
$\langle m\rangle$ is viewed as a function of repressor copy number $R$. The
second equality in~\ref{eq:fc_def} follows from assuming that the translation
efficiency, i.e., the number of proteins translated per mRNA, is the same in
both conditions. In other words, we assume that mean protein level is
proportional to mean mRNA level, and that the proportionality constant is the
same in both conditions and therefore cancels out in the ratio. This is
reasonable since the cells in the two conditions are identical except for the
presence of the transcription factor, and it is difficult to imagine the
transcription factor having any direct effect on translation.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{\textbf{An overview of the simple repression motif at the level of
means.} (A) depicts the qualitative biological picture of simple repression. (B)
shows a variety of possible mathematized cartoons of simple repression, along
with the effective parameter $\rho$ which subsumes all regulatory details of the
architecture excluding the repressor. (C) shows the ``master curve'' to which
all cartoons in (B) collapse.}
\label{fig1:means_cartoons}
\end{figure}

@@

In principle, if $\Delta F_R$ can be measured to sufficient precision, then
deviations from $\rho=1$ become a testable matter of experiment. In practice, it
is probably unrealistic to measure repressor rates $k_R^+$ or $k_R^-$ or
fold-changes in expression levels (and hence $\Delta\epsilon_R$) precisely
enough to detect the expected tiny deviations from $\rho=1$. We can estimate the
requisite precision in $\Delta F_R$ to resolve a given $\Delta\rho$ by noting,
since $\rho\approx1$, that $\log(1+\Delta\rho)\approx \Delta\rho$, so
$\Delta(\Delta F_R) \approx \Delta\rho$. Suppose we are lucky and $\Delta\rho$
happens to be $\sim0.1$, on the high end of our range estimated above. A
determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of barely 0.1 was
achieved in the excellent measurements of~\cite{Razo-Mejia2018}, so this
requires a very difficult determination of $\Delta F_R$ for a very crude
determination of $\rho$, which suggests, to put it lightly, this is not a
promising path to pursue experimentally. It is doubtful that inference of
repressor kinetic rates would be any easier.\footnote{\mmnote{In fact the
uncertainties quoted in Hammar et.\ al.~\cite{Hammar2014} from Elf's group are
small enough to do this calculation, but HJ \& I think their error bars are
insanely optimistic. Not sure I want to blatantly call them out on it though??}}

@@
